Concurrent and Distributed Programming in Harmony

### 2022-01-13 paper/paper.tex
Latest paper version

\emph{Harmony} is a Python-like programming language designed to teach
concurrent and distributed programming to students who do not have a
background in formal methods.  Harmony programs are model checked.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, trying to multiply two lists
with one another, and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\end{compactitem}

@@

Harmony programs compile to a virtual machine.  The Harmony virtual
machine is formally specified using TLA+, allowing Harmony programs
to run using TLC, the model checker that comes with the TLA+
toolbox.  Harmony has its own custom concurrent model checker that
scales significantly better for Harmony programs.

@@

@@

\paragraph{Stop/Go.}  \texttt{Stop} saves the current context (with
its program counter incremented) in a specified shared location (usually the
tail of a queue) and removes the context from the context bag and active set.
If the context was atomic (which is usually the case), the active set is
replaced with the domain of the context bag.  Even though its atomic counter
is not zero, the stopped context is not considered atomic at this time, but
will resume being atomic once it is scheduled to run.

@@

PlusCal and TLA+, Promela and Spin, DSLABS (UW), Java Pathfinder, Zing,
CBMC/JBMC.

>>>>

\emph{Harmony} is a Python-like programming language designed to teach
concurrent and distributed programming to students who do not have a
background in formal methods.  Harmony programs are model checked~\cite{CES86}.
Without any special formalism, Harmony can diagnose a large variety
of problems:
\begin{compactitem}
\item ``Silly errors'': accessing non-existent variables,
divide-by-zero, out-of-bounds array access, trying to multiply two lists
with one another, and many more.
\item Safety violations due to assertions failing, such as an assertion to ensure mutual exclusion.
\item Liveness violations (deadlock).
\item Faulty behaviors: implementation has behaviors not seen with specification, such as violations of linearizability~\cite{HW90}.
\item Incomplete behaviors: specification has behaviors not seen with implementation, such as multiple readers not able to get a reader/writer lock simultaneously.
\item Data races: variables accessed without a lock.
\item Busy Waiting: waiting for a condition without blocking.
\end{compactitem}

@@

Harmony programs compile to a virtual machine.  The Harmony virtual
machine is formally specified using TLA+~\cite{Lamport02}, allowing
Harmony programs to run using TLC, the model checker that comes
with the TLA+ toolbox.  Harmony has its own custom concurrent model
checker that scales significantly better for Harmony programs.

@@

Harmony's unique combination of features are a familiar and easy
to learn high-level programming language, the ability to concisely
specify an interface, the ability to compare behaviors of specification
and implementation, the ability to modularly model check programs,
and support for continuations and interrupts.  Harmony finds various common
problems automatically without a complicated property specification
language, such as Harmony's ability to find instances of undesirable
busy waiting.

@@

\paragraph{Stop/Go.}  Harmony supports continuations as a first-class
construct.  To enable this, \texttt{Stop} saves the current context (with
its program counter incremented) in a specified shared location (usually the
tail of a queue) and removes the context from the context bag and active set.
If the context was atomic (which is usually the case), the active set is
replaced with the domain of the context bag.  Even though its atomic counter
is not zero, the stopped context is not considered atomic at this time, but
will resume being atomic once it is scheduled to run.

@@

Harmony is preceded by many other model checkers and accompanying
languages.  This section mentions some of Harmony's strongest influences.

PlusCal~\cite{Lamport09} is an imperative programming language that,
after translation to TLA+~\cite{Lamport02} (a formal verification
language), can be run through the TLC model checker.  Harmony also
supports compilation to TLA+.  While similar in spirit to Harmony,
Harmony first compiles to a virtual machine that is close to CPUs
like x86, ARM, and RISC-V, using a stack and common control flow
techniques.  The resulting HVM has a TLA+ specification.  PlusCal,
instead, \emph{trans\-piles} to TLA+.  Informally, the gap between
PlusCal programs and their TLA+ representations is much smaller
than the gap between Harmony programs and their TLA+ representations,
making PlusCal more amenable to formal verification.  On the other
hand, working effectively with PlusCal does require significant
knowledge of TLA+, increasing the learning curve.

Promela~\cite{SPIN} is a programming language that serves as input
to the SPIN model checker.  Promela has many of the same features
as Harmony and also includes explicit message channels.  Promela
does not have a call stack and does not support procedure calls,
making it hard to do modular development and verification.

Zing~\cite{Zing} is a modeling language that, like Harmony, compiles to an
intermediate language that resembles a conventional processor,
including a call stack and dynamically allocated memory.  Like
Harmony it supports procedure calls and concurrency.  Zing goes
further in that it also support object-orientation.  Zing is
mostly used as an intermediate language, transpiled from sources
written in languages such as C\#.

Java Pathfinder~\cite{Pathfinder} and JBMC~\cite{JBMC} are model
checkers for Java bytecode, and,
like Charm, can diagnose data races and deadlock in concurrent
programs.  Because they operate at the level of Java bytecode, they
have no support for atomic code blocks, which allows Harmony to be
used both as a specification and an implementation language.
Particularly the Harmony <{atomically when}> statement is very
useful as a specification construct, for example to specify new
synchronization primitives.

DSLabs~\cite{MWA19} is an educational framework for model checking distributed
systems.  Distributed algorithms are specified in Java in terms of
nodes, messages, and timers in an event-driven style.  DSLabs does not
support model checking sequential or concurrent programs.  Java
also has the disadvantage that it is a verbose language.  Just the
boilerplate for the Paxos project in DSLabs is well over 100 lines
of Java code.  The entire Paxos implementation in Harmony is less
than 50 lines.
